A Quantitative Genetic Linked Selection Model of the Negative Selection Process in Humans

### 2023-07-21 manuscript/manuscript.tex
ms updates

    \caption{}

@@

We find that our predicted substitution rates are qualitatively consistent with
the observed divergence along the human lineage for all features except
PhastCons (Figure \ref{fig:figure-7}). As expected, the predicted substitution
rates in features unlikely to be under selective constraint (introns and UTRs,
and the ``other" feature) are very close to the mutation rate. Throughout, we
report our substitution rates as a percent relative to the total mutation rate,
$\mu$ (here fixed to $1.5 \times 10^{-8}$). In our Feature Priority model,
coding sequences are predicted to have a substitution rate of 41.20\% of the
mutation rate, introns and UTRs 94.71\%, coding sequences 0\%, and the ``other"
feature 99.98\%. For comparison, the substitution rates along the human lineage
(as a proportion to the substitution rate in putatively neutral regions) are
74.15\% in UTRs, 92.44\% in introns, 50.96\% in coding sequences, and 49.56\%
in PhastCons regions. The large discrepancy between predicted and observed
PhastCons substitution rates is driven by our DFE estimates suggesting that the
bulk of mass is on selection coefficients greater than $10^{-3}$, which have no
chance of fixation in a population of $N_e \approx 10,000$. We note that our
DFE estimates are qualitatively similar to those inferred using the classic BGS
model; thus this DFE-substitution rate contradiction is not unique to our
estimates, and could indicate a potential model misspecfication problem.

@@

Linked selection processes alter genetic diversity on multiple scales, creating
patterns based on their strength relative to the stochasticity of coalescence
times and mutations. Sweeps typically leave signals on shorter genomic scales
(i.e. kilobases; \cite{Akey2004-zg}) which combined with the paucity of
adaptive substitutions in humans suggests positive selection is generating the
apparent large, megabase-scale linked selection signal
\parencite{Hernandez2011-gs,Murphy2022-sj}. By contrast, there is reasonable
evidence that negative selection against new mutations occurring in
highly-conserved sequences is generating large-scale chromosomal patterns of
diversity.

Population genetic theory has been productively applied to fit this linked
selection signal, producing plausible estimates of the strength of selection at
these conserved regions and mutations rates consistent with more precise
pedigree-based estimates. However, these models rely on the canonical
background selection model, which due to its assumption that mutations are
sufficiently deleterious they cannot fix, has constrained statistical methods
to only fit background selection models in regions \emph{a priori} under
constraint. In reality, the relationship between diversity levels and the
strength of selection at deleterious mutations is U-shaped, which implies there
could be more uncertainty that previously appreciated in the distribution of
weak and strongly deleterious mutations.

We sought to address this uncertainty in this work, by leveraging a different
class of linked selection models based on modeling the equilibrium fitness
variance \parencite{Santiago2016-mu}. This model is more general than classic
background selection theory; the background selection model is analytically a
special case when the deleterious substitution rate is zero. Moreover, the SC16
model explains equilibrium fitness variance across all selection coefficients
by jointly predicting another central quantity in evolutionary genomics: the
deleterious substitution rate in conserved features. This offers a further way
to confirm or falsify that our understanding of the negative selection process
is accordance with other observed quantities.

However, while our improved negative selection model leads to nearly identical
parameter estimates as previous working employing classic BGS theory, we find a
fundamental contradiction in the implied rate of deleterious substitution and
observed divergence in conserved regions. Unfortunately, this suggests a
potentially serious model misspecfication or non-identifiability issue. Since
the rate of substitutions is determined by the fraction of DFE mass below $2N_e
s \le 1$, the consistency between our DFE estimates and those from
\textcite{Murphy2022-sj} suggests this discrepancy is shared across models of
negative selection.

One clue towards solving this contradiction is given by our simulation work.
While our linked selection model accurately predicts the substitution rates,
fitness variance, and diversity for $2N_e s > 1$ and $2N_e s \ll 1$, there's a
nominal error around the drift boundary of $2N_e s \approx 1$. This region of
parameter space is at a delicate balance between stochasticity dominating
allele frequency dynamics (for $2N_e s \ll 1$) and having little influence
($2N_e s > 1$). Our implementation of the SC16 model in essence only considers
the reduction due to selection \emph{within} the segment and not neighboring
regions. Thus, like classic BGS methods, our model does not consider selective
interference due to neighboring regions under selective constraint. Solving the
SC16 equations using a locally-rescaled population size of $B(x) N$ rather than
$N$ are in effect approximating the reduction in effective population due to
selective interference. Intuitively, this is based on the notion that if a
neutral allele experiences an effective population size of $B(x)N$, so too
should a selected allele. We find in simulations this reduces error around $2
N_e s \approx 1$. Unfortunately, employing this in our composite-likelihood
method would create some circularity, since the local $B(x)$ experienced at a
conserved segment includes the reduction due to the segment itself. There are
not computationally feasible ways to avoid this issue in the current
implementation.

Out of curiosity, we experimented with fitting human diversity data, using the
predicted $\widehat{B}(x)$ for local-rescaling, and rerunning our
composite-likelihood framework (see Supplementary Mathematical XXX). While the
circularity issue means these results should be interpreted with caution, we
generally find these models fit nearly as well

it creates a bit of circularity.

draft limits discussion - we were heavily constrained by simulations / regions
etc

the PhastCons Priority Model is very close to CADD 6\%, and interpretable,
similiar mutation rates. This

weak selection remains elusive

UTR rates

- parametric DFEs are trash

- the $pi_0$ problem

Ipsita Agarwal article

what time scale is the average mutation being lost at?

Regularization / overfitting

It is worth considering why this model performs well for models of negative
selection. The fitness variance equation, $V(x) = (U-2R)s$ tells us that under
models, the long-run variance is set by the balance of mutation and fixation.
Whether a particular basepair's behavior follows this equation depends upon
whether the long-run average is close to ``typical" for that basepair at a
given time. Good things fix so quickly, that we're left with a constant
churning of new variation

- need for coalescent simulations with reduced $N_e$ along the genome.

- we need to talk about Fisher's Fundamental Theorem and how this model
connects fitness variation to absolute changes in fitness.

Measuring selection in the Genome Versus Finding Selection in the Genome

\section*{Methods}
% (the method would support a new
% ``interaction" type, but this greatly increases the number of parameters).

@@

For all of our main analyses, we used the recombination map from
\textcite{Halldorsson2019-ey} estimated from a trio-based design to avoid
circularity that could occur by using LD-based maps. We use Ensembl gene
annotation \parencite{Cunningham2022-vk}, a special CADD Score dataset with
McVicker B scores removed (to avoid circularity;
\cite{Kircher2014-bv,Rentzsch2019-lr}), and PhastCons regions
\parencite{Siepel2005-wh}.

>>>>

    \caption{Predicted versus observed divergence along the human lineage.
        Black points are the PhyloFit divergence rate estimates per features
        (on x-axis). Line ranges are the implied divergences across a range of
        human-chimpanzee divergence times of 6-12 Mya (using a generate time of
        29 years). We show the predicted divergences for our Feature
        (turquoise) and PhastCons priority (green) annotation models.
        Additionally, we show the predicted PhastCons region divergences when
    local-rescaling is applied (we omit other locally-rescaled predictions
since these to not differ substantially).}

@@

We find that our predicted substitution rates are qualitatively consistent with
the observed divergence along the human lineage for all features except
PhastCons (Figure \ref{fig:figure-6}). As expected, the predicted substitution
rates in features unlikely to be under selective constraint (introns and UTRs,
and the ``other" feature) are very close to the mutation rate. Throughout, we
report our substitution rates as a percent relative to the total mutation rate,
$\mu$ (here fixed to $1.5 \times 10^{-8}$). In our Feature Priority model,
coding sequences are predicted to have a substitution rate of 41.20\% of the
mutation rate, introns and UTRs 94.71\%, coding sequences 0\%, and the ``other"
feature 99.98\%. For comparison, the substitution rates along the human lineage
(as a proportion to the substitution rate in putatively neutral regions) are
74.15\% in UTRs, 92.44\% in introns, 50.96\% in coding sequences, and 49.56\%
in PhastCons regions. The large discrepancy between predicted and observed
PhastCons substitution rates is driven by our DFE estimates suggesting that the
bulk of mass is on selection coefficients greater than $10^{-3}$, which have no
chance of fixation in a population of $N_e \approx 10,000$. We note that our
DFE estimates are qualitatively similar to those inferred using the classic BGS
model; thus this DFE-substitution rate contradiction is not unique to our
estimates, and could indicate a potential model misspecfication problem.

@@

Linked selection processes alter genetic diversity on multiple scales, creating
patterns based on their strength relative to the stochasticity of coalescence
times and mutations. Sweeps typically leave signals on shorter genomic scales
(i.e. kilobases; \cite{Akey2004-zg}) which combined with the paucity of
adaptive substitutions in humans suggests positive selection is not generating
the apparent large, megabase-scale linked selection signal
\parencite{Hernandez2011-gs,Murphy2022-sj}. By contrast, there is reasonable
evidence that negative selection against new mutations occurring in
highly-conserved sequences is generating large-scale chromosomal patterns of
diversity.

Population genetic theory has been productively applied to fit this linked
selection signal, producing plausible estimates of the strength of selection at
these conserved regions and mutations rates consistent with more precise
pedigree-based estimates. However, these models rely on the canonical
background selection model, which due to its assumption that mutations are
sufficiently deleterious they cannot fix, has constrained statistical methods
to only fit background selection models in regions \emph{a priori} under
constraint. In reality, the relationship between diversity levels and the
strength of selection at deleterious mutations is U-shaped, which implies there
could be more uncertainty than previously appreciated in the distribution of
weak and strongly deleterious mutations.

We sought to address this uncertainty in this work, by leveraging a different
class of linked selection models based on modeling the equilibrium fitness
variance \parencite{Santiago2016-mu}. At its core, the theory underlying this
model states that the linked selection signal is a function of how the fitness
variance is distributed along the recombination map of the genome
\parencite{Santiago1998-bs}. We fit a specific process model for this fitness
variance that supposes all fitness variance is the result of selection against
new deleterious mutations \parencite{Santiago2016-mu}. Unlike classic
background selection theory, the SC16 model explains equilibrium fitness
variance across all selection coefficients by jointly predicting another
central quantity in evolutionary genomics: the deleterious substitution rate in
conserved features. This model is more general than classic background
selection theory; background selection theory is analytically a special case
when the deleterious substitution rate is zero.

Generally, our method lead to better fits to data that those based on classic
BGS, as measured by predicted out-sample diversity. Furthermore, unlike
BGS-based methods, our model is capable of fitting weak selection; we find
evidence of this since we can robustly estimate the DFE for weakly constrained
features like introns and UTRs. Yet even with the possibility of fitting weak
selection, our main model fits are in agreement with the prior results from
BGS-based methods. Both mutation rate estimates and DFE estimates are
consistent with prior work. This, at first glance, suggests further
confirmation that negative selection against new deleterious mutations is the
dominant mode of linked selection in the genome.

Since the SC16 theory calculates fitness variance by jointly predicting the
deleterious substitution rate, this provides another route to verify our main
findings. We find that predicted substitution rates for highly-conserved
PhastCons features are in strong disagreement with observed substitution rates
along the human lineage: the DFE-substitution rate contradiction. This is a
consequence of our DFE estimate for PhastCons regions containing little mass
over weakly deleterious and neutral selection coefficients --- a characteristic
of DFE estimates from other work too \parencite{Murphy2022-sj}. One clue to
this discrepancy is that our simulation results indicate an appreciable level
of error between observed reductions and the B' maps in the weak selection
domain of $2N_e s \approx 1$.

We hypothesized that this could be because in this domain, a segment under
selective constraint experiences a local draft-effective population size of
$B(x)N$, and not just $N$. This local draft-effective population size is
induced by selection at other segments that is not being taken into account by
classic BGS theory or our standard model. When we experimented by fitting our
model and then using the predicted reduction map to local-rescale $N_e$ to the
draft-effective population size $N_d = \widehat{B}(x)$, we find the
disagreement between predicted and observed substitution rates disappears. This
is expected, since locally-rescaled fits result in DFE estimates to have an
appreciable mass on weakly deleterious selection coefficients. While this
brings substitution rates into agreement, we encounter the same problem found
by \textcite{McVicker2009-ax}: the estimated mutation rate is too high. We
emphasize that although there are several caveats with the local-rescaling
approach, in practice it is altering the B' maps only slightly in the domain of
$2N_e s \approx 1$. Even if the local-rescaling approximation to selective
interference is incorrect, these results suggest parameter estimates are
extremely sensitivity to how accurate the B' maps are in this domain.

Unfortunately, this work opens up more questions about the nature of selection
in humans that it answers. The implausibly high estimate of mutation rates
could suggest that negative selection is not the only source of fitness
variance generating the genome-wide linked selection signal in humans. Positive
selection on polygenic variation along the genome could be another source of
fitness variance contributing to the weak selection signal. Alternatively,
negative selection could be the main source of fitness variance, but the
complexity of selective interference may be poorly approximated by the
local-rescaling approach. Overall, understanding how the sources of human
fitness variance shape the striking patterns of diversity along the human
genome still remains an open riddle.

\section*{Methods}

@@

For all of our main analyses, we used the recombination map from
\textcite{Halldorsson2019-ey} estimated from a trio-based design to avoid
circularity that could occur by using LD-based maps. We use Ensembl gene
annotation \parencite{Cunningham2022-vk}, a special CADD Score dataset with
McVicker B scores removed (to avoid circularity;
\cite{Kircher2014-bv,Rentzsch2019-lr}), and PhastCons regions
\parencite{Siepel2005-wh}. We did not account for mutation rate heterogeneity
along the genome, since this would require using divergence-based estimates of
local mutation rates that would introduce circularity when we predict
divergence rates under our model.
